The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-15 talk/vmil2012/paper.tex
some protected whitespace

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase guards are
inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after
optimizing the traces generated for the different benchmarks used in this paper.
This makes guards one of the most common types of operations. As this paper will show, many of these guards
fail rarely or not all during execution.

@@

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend related to recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

After optimization the resulting trace is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the backend map and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

>>>>

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section~\ref{sec:Resume Data}). During the recording phase guards are
inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after
optimizing the traces generated for the different benchmarks used in this paper.
This makes guards one of the most common types of operations. As this paper will show, many of these guards
fail rarely or not all during execution.

@@

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend related to recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section~\ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

After optimization the resulting trace is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing the overhead of the guard.
Figure~\ref{fig:trace-compiled} shows how the \texttt{int\_eq} operation
followed by a \texttt{guard\_false} from the trace in Figure~\ref{fig:trace-log} are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard~\#1 jumps to the trampoline, loads the backend map and
then calls the bailout handler, whereas Guard~\#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)
